---
title: "Assignment_2"
author: "James Gramajo"
date: "2025-02-25"
output: html_document
---
Part 1
1.1 Functions
```{r}
print_message <- function() {
  print("Hello world!")
}
print_message()
```

1.2 Mandatory parameters
```{r}
hello_function <- function(name, age) {
  print(paste("Hola: ",name, " Age: ",age))
}
hello_function("James", 24)  #parameters are required for printing
```

1.3 Parameters with default value
```{r}
sum_function <- function(n1, n2 = 100) {
  n1+n2
}
print(sum_function(50))     # Use 100 by default
```

1.4 List parameters
```{r}
parameters_list <- function(lst) {
  list(
    suma = sum(lst)
  )
  
}
print(parameters_list(c(5,5,10,10,10))) ##add the parameters
```

1.5 Function passing parameters
```{r}
apply_function <- function(f, x) {
  return(f(x))
}
apply_function(sqrt, 16)  #the square root function is passed as a parameter
```

1.6 Function return 
```{r}
return_with_function <- function(n1,n2,n3) {
  return(n1+n2+n3)
}
result <- return_with_function(2,2,6)  # Devuelve 10
print(result)
```

1.7 Local and global variables
```{r}
global_var <- 500

function_global_local <- function() {
  local_var <- 100
  return(global_var + local_var)
}
print(function_global_local())  
```

1.8 Functions that return multiple values
```{r}
two_value <- function(a,b) {
  return(list(addition = a + b, subtraction = a - b))
}
print(two_value(10, 5)  )
# Returns two values, the first the sum of the two numbers and 
# the second the subtraction of the two numbers
```


PART 2
2.1 Matrix Manipulation
2.1.1 Create a 4x4 matrix with random numbers between 1 and 100.
```{r}
matrix4 <- matrix(sample(1:100, 16, replace = TRUE), nrow = 4, ncol = 4)
```

2.1.2 Name the rows as "A", "B", "C", "D" and the columns as "X", "Y", "Z", "W".
```{r}
rownames(matrix4) <- LETTERS[1:4]
colnames(matrix4) <- c("X", "Y", "Z", "W")

```
2.1.3 Calculate the mean per row and per column using rowMeans() and colMeans().
```{r}
row_mean <- rowMeans(matrix4)
colum_mean <- colMeans(matrix4)

```

2.1.4 Add a new column to the matrix with the mean of each row using cbind().
```{r}
matrix4_final <- matrix4
matrix4_final <- cbind(matrix4_final, Media_Fila = row_mean)
```

2.1.5 Add a new row with the mean of each column using rbind().
```{r}
matrix4_final <- rbind(matrix4_final, Media_Col = c(colum_mean, NA))
```
 
2.1.6 Display the resulting matrix.
```{r}
print(matrix4_final)
```

2.2 Creating and Manipulating Dataframes
2.2.1 Create a dataframe with information on 5 employees, with the following columns:
- ID (a unique number for each employee).
- Name (employee's name).
- Salary (random number between 2000 and 8000).
- Department (randomly assigns between "Sales", "IT", "HR", "Marketing").
- Calculate the average salary of the employees using mean().
- Filter and display employees whose salary is higher than the average salary.
- Add a new Bonus column, which is 10% of salary only if the employee belongs to "Sales" or "Marketing".
```{r}
employees <- data.frame(
  ID = 1:5, 
  Nombre = c("Cinthia", "Natalia", "Elizabeth", "Jazlyne", "Jarissa"),
  Salario = round(runif(5, 2000, 8000)),
  Departamento = sample(c("Ventas", "IT", "Recursos Humanos", "Marketing"), 5, replace = TRUE)
)
```


2.2.2 Calculate the average salary of employees using mean().
```{r}
average_salary  <- mean(employees$Salario)
```


2.2.3 Filter and display employees whose salary is higher than the average salary.
```{r}
filtered_employees <- subset(employees, Salario > average_salary)
print(filtered_employees)
```

2.2.4 Add a new Bonus column that is 10% of salary only if the employee belongs to "Sales" or "Marketing."
```{r}
employees$Bono <- ifelse(employees$Departamento %in% c("Ventas", "Marketing"),
                         employees$Salario * 0.10, 0)
print(employees)
```



2.3 Using Lists
2.3.1 Create a list in R containing the matrix from section 1 and the data frame from section 2.
2.3.2 Display the data in the list in a structured manner.
```{r}
# Using lists

# Create a list containing the matrix and the data frame
list_data <- list(
  matriz = matrix4_final,  # Matrix of section 1
  employees = employees  # Dataframe from section 2
)

# Display list data in a structured manner
print(list_data)

```
2.4 Defining Custom Functions
2.4.1 Write a function called adjust_salary() that takes a data frame and an adjustment value, and increases all employees' salaries by that percentage.

2.4.2 Test the function with a 5% adjustment and display the result.

```{r}
# Defining custom functions
adjust_salary <- function(data_frame, ajuste) {
  data_frame$Salario <- data_frame$Salario * ((100+ajuste)/100)
  return(data_frame)
}
# Call the function with a 5% setting
employees <- adjust_salary(employees, 5)
print(employees)

```
2.5 Functions with Parameters and Data Filtering
2.5.1 Define a function, filter_department(), that takes a data frame and a department name as parameters and returns only the employees in that department.
2.5.2 Execute the function and display the employees in "IT."

```{r}
filter_department <- function(dataframe, department) {
  return(dataframe[dataframe$Departamento == department, ])
}

# Call the function and display the IT employees

IT_department_jobs <- filter_department(employees, "IT")
print(IT_department_jobs)

```

